---
title: "Getting started with utilscidados"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with utilscidados}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(utilscidados)
```

## What this package is for

`utilscidados` collects the small helpers we reach for at Insper
Cidades when preparing data for two specific destinations:

1.  **Insper's Dataverse.** Datasets published there must be shipped
    in several file formats with predictable naming and accompanied
    by column-level documentation. `export_table()` /
    `export_shapefile()` produce the full Dataverse-compliant set in
    one call (`extension = "dataverse"`), while
    `build_documentation()` and `create_documentation_index()` build
    the documentation tables.

2.  **Mapbox Studio (GeoPortal).** Spatial layers are uploaded as
    GeoJSON or GeoPackage in WGS84. `export_shapefile()` reprojects
    the input, sanitises the file name, and writes both formats so
    the layer is ready to upload without per-dataset wrangling.

All exporters share consistent defaults, overwrite protection, and
`cli`-styled progress messages.

The chunks below are not evaluated when the vignette is built (they
would litter `tempdir()` with files); copy them into a session if you
want to try them.

## Exporting a tabular dataset

`export_table()` writes a data frame to one or more file formats in a
common output directory:

```{r}
out <- tempdir()

# Single format
export_table(
  iris,
  out_dir   = out,
  file_name = "iris",
  extension = "parquet"
)

# Multiple formats at once
export_table(
  iris,
  out_dir   = out,
  file_name = "iris",
  extension = c("csv", "parquet", "rds"),
  overwrite = TRUE
)
```

`extension = "all"` writes every supported format; `extension =
"dataverse"` writes the formats accepted by Dataverse (csv, xlsx,
parquet).

CSV output is gzip-compressed for data frames above 2 million rows
(file extension becomes `.csv.gz`). XLSX output above the Excel row
limit still attempts to write, but emits a warning.

## Exporting a spatial dataset

`export_shapefile()` is the spatial counterpart. The argument shape
matches `export_table()`:

```{r}
nc <- sf::st_read(system.file("shape/nc.shp", package = "sf"))

export_shapefile(
  nc,
  out_dir   = out,
  file_name = "north_carolina",
  extension = "dataverse"
)
```

Notes:

* GeoJSON is reprojected to WGS84 (EPSG:4326).
* Shapefile output is placed in a subdirectory named after the file,
  because a shapefile is really a set of sidecar files (`.shp`,
  `.shx`, `.dbf`, ...).
* GeoParquet is written via `sfarrow::st_write_parquet()`.

## File name sanitization

Both `export_table()` and `export_shapefile()` lower-case the
`file_name` and replace runs of non-alphanumerics with `_`. This means
the same input produces a filesystem-safe name on every platform:

```{r}
# These all produce the same base file name: "north_carolina_2024"
export_table(iris, file_name = "North Carolina 2024", out_dir = out)
export_table(iris, file_name = "north-carolina-2024", out_dir = out)
export_table(iris, file_name = "North_Carolina/2024", out_dir = out)
```

## Multi-sheet Excel workbooks

When you want a single `.xlsx` file with one sheet per data frame,
reach for `export_tables_to_excel()`:

```{r}
tables <- list(
  iris   = iris,
  mtcars = mtcars
)

export_tables_to_excel(
  tables    = tables,
  file_name = file.path(out, "example.xlsx")
)
```

Names of the input list become sheet names, truncated to 31 characters
(Excel's limit) and made unique if needed. Headers are bold; column
widths are auto-sized.

## Building a data dictionary

For published datasets we usually ship a column-level data dictionary
alongside the data. `build_documentation()` profiles each column
(type, sample values, missingness) and merges in any manual
descriptions you provide:

```{r}
descs <- tibble::tibble(
  col_names   = c("Sepal.Length", "Species"),
  description = c("Sepal length in cm", "Iris species")
)

doc <- build_documentation(iris, tbl_description = descs)
doc
```

The resulting tibble has the Portuguese-language column labels we use
in published documentation (`Nome da Coluna`, `Tipo da Coluna`,
`Valores`, `Descrição`, `Percentual de Obs. Ausentes`,
`Percentual de Preenchimento`).

`build_documentation()` works on plain data frames and on `sf`
objects; if you pass `sf`, geometry is dropped before profiling.

When you ship more than one table, build a small index too:

```{r}
create_documentation_index(
  name_file  = c("dados_2023.csv", "dados_2024.csv"),
  name_table = c("dados_2023", "dados_2024")
)
```

A typical publication workflow then becomes:

```{r}
docs <- lapply(list(iris, mtcars), build_documentation)
names(docs) <- c("iris", "mtcars")

# Combine with the file/table index as the first sheet
docs <- c(
  list(`indice` = create_documentation_index(
    name_file  = c("iris.csv", "mtcars.csv"),
    name_table = c("iris", "mtcars")
  )),
  docs
)

export_tables_to_excel(docs, file.path(out, "documentation.xlsx"))
```
