---
title: "Getting started with utilscidados"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with utilscidados}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(utilscidados)
```

## What this package is for

`utilscidados` collects the small helpers we use at Insper Cidades
when preparing data for two destinations:

1.  **Insper's Dataverse.** Datasets published there must ship in
    several file formats with predictable naming and be accompanied
    by column-level documentation. `export_table()` /
    `export_shapefile()` produce the Dataverse-compliant set in
    one call (`extension = "dataverse"`), while
    `build_documentation()` and `create_documentation_index()` build
    the documentation tables.

2.  **Mapbox Studio (GeoPortal).** Spatial layers are uploaded as
    GeoJSON or GeoPackage in WGS84. `export_shapefile()` reprojects
    the input, sanitises the file name, and writes both formats so
    the layer is ready to upload without manual preparation.

The chunks below are not evaluated when the vignette is built; copy
them into a session if you want to try them.

## Exporting a tabular dataset

`export_table()` writes a data frame to one or more file formats in a
common output directory:

```{r}
out <- tempdir()

# Single format
export_table(iris, "iris", out_dir = out, extension = "parquet")

# Multiple formats at once
export_table(
  iris,
  "iris",
  out_dir = out,
  extension = c("csv", "parquet", "rds"),
  overwrite = TRUE
)
```

`extension = "all"` writes every supported format; `extension =
"dataverse"` writes the formats accepted by Dataverse (csv, xlsx,
parquet).

CSV output is gzip-compressed for data frames above 2 million rows
(file extension becomes `.csv.gz`). XLSX output above the Excel row
limit still attempts to write, but emits a warning.

## Exporting a spatial dataset

`export_shapefile()` is the spatial counterpart. It takes the same
arguments as `export_table()`:

```{r}
nc <- sf::st_read(system.file("shape/nc.shp", package = "sf"))

export_shapefile(nc, "north_carolina", out_dir = out, extension = "dataverse")
```

Notes:

* GeoJSON is reprojected to WGS84 (EPSG:4326).
* Shapefile output is placed in a subdirectory named after the file,
  because a shapefile is really a set of sidecar files (`.shp`,
  `.shx`, `.dbf`, ...).
* GeoParquet is written via `sfarrow::st_write_parquet()`.

## Uploading a layer to Mapbox Studio (GeoPortal)

For layers that need to be served from GeoPortal, `mapbox_upload()`
sends the data straight to Mapbox as a tileset. It accepts either an
`sf` object or a path, can simplify the geometry on the way, and can
wait until Mapbox finishes processing the upload.

```{r}
# Requires the suggested `mapboxapi` package (and `rmapshaper` if
# `simplify` is not FALSE). Reads MAPBOX_SECRET_TOKEN from env by
# default.
mapbox_upload(
  nc,
  username = "insper_cidados",
  tileset_id = "nc_counties",
  simplify = 0.1, # rmapshaper::ms_simplify(keep = 0.1)
  wait = TRUE
)
```

## File name sanitisation

`export_table()`, `export_shapefile()`, and `mapbox_upload()` all
lower-case the supplied name and replace runs of non-alphanumerics
with `_`. The same input produces a filesystem-safe / Mapbox-safe
name on every platform:

```{r}
# These all produce the same base name: "north_carolina_2024"
export_table(iris, "North Carolina 2024", out_dir = out)
export_table(iris, "north-carolina-2024", out_dir = out)
export_table(iris, "North_Carolina/2024", out_dir = out)
```

## Multi-sheet Excel workbooks

When you want a single `.xlsx` file with one sheet per data frame,
reach for `export_tables_to_excel()`:

```{r}
tables <- list(
  iris = iris,
  mtcars = mtcars
)

export_tables_to_excel(
  tables = tables,
  file_name = file.path(out, "example.xlsx")
)
```

Names of the input list become sheet names, truncated to 31 characters
(Excel's limit) and made unique if needed. Headers are bold; column
widths are auto-sized.

## Building a data dictionary

For published datasets we usually ship a column-level data dictionary
alongside the data. `build_documentation()` profiles each column
(type, sample values, missingness) and merges in any manual
descriptions you provide:

```{r}
descs <- tibble::tibble(
  col_names = c("Sepal.Length", "Species"),
  description = c("Sepal length in cm", "Iris species")
)

doc <- build_documentation(iris, tbl_description = descs)
doc
```

The resulting tibble has the Portuguese-language column labels we use
in published documentation (`Nome da Coluna`, `Tipo da Coluna`,
`Valores`, `Descrição`, `Percentual de Obs. Ausentes`,
`Percentual de Preenchimento`).

`build_documentation()` works on plain data frames and on `sf`
objects; if you pass `sf`, geometry is dropped before profiling.

When you ship more than one table, build a small index too:

```{r}
create_documentation_index(
  name_file = c("dados_2023.csv", "dados_2024.csv"),
  name_table = c("dados_2023", "dados_2024")
)
```

In a typical publication workflow, you build the per-table docs and
combine them with the index into one workbook:

```{r}
docs <- lapply(list(iris, mtcars), build_documentation)
names(docs) <- c("iris", "mtcars")

# Combine with the file/table index as the first sheet
docs <- c(
  list(
    `indice` = create_documentation_index(
      name_file = c("iris.csv", "mtcars.csv"),
      name_table = c("iris", "mtcars")
    )
  ),
  docs
)

export_tables_to_excel(docs, file.path(out, "documentation.xlsx"))
```
